---
title: "gapminder"
author: "Ken"
date: "3/13/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r}
library(tidyverse, echo=FALSE)

```
Read some data.
```{r}
gapminder <- read.csv("https://raw.githubusercontent.com/jules32/2017-11-30-MBARI/gh-pages/data/gapminder.csv")
```
Show the first few rows or the last few rows and the structure of the dataset 
```{r}
head(gapminder)
tail(gapminder)
str(gapminder)
```

Look at some summaries to help clean data
```{r}
#install.packages("skimr)
library(skimr)
skim(gapminder)
```
extract and filter data
```{r}
filter(gapminder, country=="Switzerland" )
#same as
gapminder[gapminder$country=="Switzerland",]
filter(gapminder,lifeExp < 25)
filter(gapminder, lifeExp==max(lifeExp))
```
mess with columns
```{r}
dplyr::select(gapminder, country, year, pop)
gapminder_no_cont <- dplyr::select(gapminder, -continent)
filter(gapminder_no_cont,year==2007)
```
Mess with pipes - this is only in tidyverse not base R

```{r}
dplyr::select(gapminder, -continent) %>%  filter(year==2007)
```
Change some columns
```{r}
mutate(gapminder, totalGdp= pop * gdpPercap) %>%
  arrange(desc(totalGdp))
```
Try analysis on  groups

```{r}
gapminder %>%
  group_by(country) %>%
  mutate(gdp= gdpPercap*pop,max_gdp=max(gdp)) %>%
  ungroup() %>%
  filter(gdp==max_gdp)
```

summary data by continent
```{r}
gapminder %>% 
  group_by(continent) %>%
  mutate(gdp= gdpPercap*pop) %>%
  summarise(max_gdp=max(gdp),min_gdp=min(gdp))
```
joining tables
```{r}
co2 <- readr::read_csv("https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/co2.csv")
co2 %>% head()
co2 %>% dim() # 12

```
join save 
```{r}
gap_2007 <- gapminder %>%
  filter(year== 2007)
inner_join(gap_2007,co2,by="country") #%>%
  #filter(is.na(co2_2007))

```

#Tidyr
Normalising table 
```{r}
gap_wide <- readr::read_csv("https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/gapminder_wide.csv")
```

```{r}
gap_better <- gap_wide %>% 
  gather(observation,observ_value,-continent,-country) %>%
  separate(observation,into= c("obs_type","year"), sep="_",convert=TRUE)
```
try the spread function 
```{r}
unique(gap_better$obs_type)
```
```{r}
gap_better %>%
  spread(obs_type,observ_value)
```


